Improving Item-Item Similarity Estimation during Collaborative Filtering

### 2014-03-24 paper/paper.tex
small changes to paper

We model the problem probabilistically. Let $Y$ be a random variable
representing the $TrueSim$ of a randomly chosen pair of items. Let $X_n$ be a
random variable representing the calculated $Sim$ of a pair of items with $n$
common users.

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=0.5\textwidth]{twonormals.png}
	\caption{The blue curve represents $Y$, the distribution of true
    similarities between a random pair of items. The red curve represents $X|Y$,
    the observed similarity when a small number of common users exist.}
    \label{fig:two_normals}
\end{figure}

In probabilistic terms, for a calculated value $X_n$, we want to find the value
of $Y$ that most likely produced $X_n$. Thus we desire the Maximum Likelihood
Estimate of $Y | X_n$
\begin{align}
\argmax_yP(Y=y|X_n=x) &= \argmax_y\left[\frac{P(X_n=x |
Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
Solving for y:
\begin{align}
y &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(y - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

The model suggests that there is a linear correlation between $X_n$, the
similarity calculated with $n$ common users, and $Y$, the true similarity.
Moreover, the best predicted $y$ is a linear combination of the mean and the
observed similiatiry. Since $\sigma_{2,n}^2\ge0$, the slope is always $\le1$, 
meaning that the calculated distance from $\mu$ (the average similarity) 
overestimates the true distance. This makes sense, since there is a prior 
distribution of true similarity weighted around $\mu$.

The parameters $\mu$, $\sigma_{1}^2$, and $\sigma_{2,n}^2$ all can be approximated 
using the method of moments over training data where there are enough common
users that $TrueSim$ can be approximated. This method requires approximating
$\sigma_{2,n}^2$ for every value of $n$ separately. Alternatively, we can model
$\sigma_{2,n}^2$ as a function of $n$. Recall that $\sigma_{2,n}^2$ is the
noisiness of the measurement $Sim$ about $TrueSim$ and $TrueSim = lim_{n \to
\infty}Sim$. Thus $Sim$ can be thought of as a sampling $n$ users from the
infinite set used to calculate $TrueSim$. Although $Sim$ may not be linear, we
the Central Limit Theorem motivates the intuition that the variance of $Sim$
would decrease as the $sqrt(n)$. Thus we model $\sigma_{2,n}^2$ as:
\begin{align}
\sigma_{2,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}
In terms of a modified similarity function, as desired earlier, we now have:
\begin{align}
ModSim(Sim, n) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(Sim-\mu\right) + \mu
\end{align}

>>>>

We model the problem probabilistically. Let $Y$ be a random variable
representing the $TrueSim$ of a randomly chosen pair of items. Let $X_n$ be a
random variable representing the observed $Sim$ of a pair of items with $n$
common users.

@@

\begin{figure}[!htbp]
    \centering
    \includegraphics[width=0.7\textwidth]{twonormals.png}
	\caption{The blue curve represents $Y$, the distribution of true
    similarities between a random pair of items. The red curve represents $X|Y$,
    the observed similarity when a small number of common users exist.}
    \label{fig:two_normals}
\end{figure}

In probabilistic terms, for an observed value $X_n$, we want to find the value
of $Y$ that most likely produced $X_n$. Thus we desire the Maximum Likelihood
Estimate of $Y | X_n$
\begin{align}
\argmax_yP(Y=y|X_n=x) &= \argmax_y\left[\frac{P(X_n=x |
Y=y)P(Y=y)}{P(X_n=x)}\right]
\\&= \argmax_y\left[P(X_n=x|Y=y)P(Y=y)\right]
\\&= 
\argmax_y\left[\frac{1}{\sigma_{2,n}\sqrt{2\pi}}\exp{\left(\frac{-(x-y)^2}
{2\sigma_{2,n}^2}\right)}
\frac{1}{\sigma_{1}\sqrt{2\pi}}\exp{\left(\frac{-(y-\mu)^2}
{2\sigma_{1}^2}\right)}\right]
\\&= \argmin_y\left[\frac{(x-y)^2}{2\sigma_{2,n}^2} +
\frac{(y-\mu)^2}{2\sigma_{1}^2}\right]
\\&= \argmin_y\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
\end{align}
We find the exact minimum by taking the derivative with respect to $y$ and
setting it to zero:
\begin{align}
\frac{d}{dy}\left[\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y^2 - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right)y\right]
&= 2\left(\sigma_{1}^2+\sigma_{2,n}^2\right)y - 
2\left(\sigma_{1}^2x+\sigma_{2,n}^2\mu\right) 
\\&= 0
\end{align}
Solving for y:
\begin{align}
y &= \frac{\sigma_{1}^2x+\sigma_{2,n}^2\mu}{\sigma_{1}^2+\sigma_{2,n}^2}
\end{align}
which can be rewritten as:
\begin{align}
\left(y - \mu\right) &= \frac{\sigma_{1}^2}{\sigma_{1}^2+\sigma_{2,n}^2}
\left(x-\mu\right)
\end{align}

The model suggests that there is a linear correlation between $X_n$, the
observed similarity with $n$ common users, and $Y$, the true similarity.
Moreover, the best predicted $y$ is a linear combination of the mean and the
observed similiarity. Since $\sigma_{2,n}^2\ge0$, the slope is always $\le1$, 
meaning that the observed distance from $\mu$ (the average similarity) 
overestimates the true distance. This is reasonable, since there is a prior 
that expects the true similarity to be around $\mu$.

The parameters $\mu$, $\sigma_{1}^2$, and $\sigma_{2,n}^2$ can be approximated 
using the method of moments over training data where there are enough common
users that $TrueSim$ can be approximated. This method requires approximating
$\sigma_{2,n}^2$ for every value of $n$ separately. To gain a more general 
model, we can model $\sigma_{2,n}^2$ as a function of $n$. Recall that 
$\sigma_{2,n}^2$ is the noisiness of the measurement $Sim$ about $TrueSim$ and $TrueSim = lim_{n \to
\infty}Sim$. Thus $Sim$ can be thought of as a sampling $n$ users from the
infinite set used to calculate $TrueSim$. Although $Sim$ may not be linear, we
the Central Limit Theorem motivates the intuition that the variance of $Sim$
would decrease as the $sqrt(n)$. Thus we model $\sigma_{2,n}^2$ as:
\begin{align}
\sigma_{2,n}^2 = \frac{\alpha}{\sqrt{n}},
\end{align}
yielding the single parameter model:
\begin{align}
\left(Y - \mu\right) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(X_n-\mu\right)
\end{align}
In terms of a modified similarity function, as desired earlier, we now have:
\begin{align}
ModSim(Sim, n) = \frac{\sigma_{1}^2}{\sigma_{1}^2+\frac{\alpha}{\sqrt{n}}}
\left(Sim-\mu\right) + \mu
\end{align}
